---
title: "Differentiation statistics analysis using LNENs samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LCNEC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  #warning = FALSE,  # Disable warnings
)
```

```{r setup, warning = FALSE, message=FALSE}
library(PopGenBounds)
library(dplyr)
library(patchwork)
library(glue)
```

## Load data
(To do) Some description about where the data come from.
```{r load data}
path_to_data <- "C:/Users/limy/OneDrive - International Agency for Research on Cancer/Documents/NoahCollab/code/MS_panNEN_organoids/data/small_variants_CCFs"
setwd(path_to_data)

data_frame_names <- list.files(pattern = "*.tsv")       # Get all file names

file_names_3_subpop <- c("PANEC1_annotatedvariants_CCF_clonality.tsv", "LCNEC4_annotatedvariants_CCF_clonality.tsv")
file_names_2_subpop <- data_frame_names[!data_frame_names %in% file_names_3_subpop]

data_frame_list_2pop <- lapply(file_names_2_subpop, read.delim)  # Read all data frames
data_frame_list_3pop <- lapply(file_names_3_subpop, read.delim)  # Read all data frames
```
## Filter data

Let's choose two samples to demonstrate the package's performance: one containing 2 subpopulations \((K = 2)\), and another with 3 subpopulations \((K = 3)\).

We will first filter the data to exclude clonal mutations (i.e., keep only subclonal mutations).  
The data should contain one row per locus, andcolumns with:
  - allele frequencies in tumor subpopulations,  
  - predicted cluster,  
  - clonality status,  
  - and a **`Sample`** column indicating the subpopulation name (?? what is actualy there?).

You may optionally exclude non-mutated loci (i.e., all allele frequencies equal to 0) using the `filter_fixed` parameter.


```{r filter data}
LCNEC10 <- data_frame_list_2pop[[2]]
PANEC1 <- data_frame_list_3pop[[1]]

LCNEC10_clean <- filter_data(LCNEC10, K=2, filter_fixed=T)
PANEC1_clean <- filter_data(PANEC1, K=3, filter_fixed=T)
```
## Plot allele frequencies
To check data purity one may look at the distribution of allele frequencies for pairs of subpopulations. In the data we use, clustering of data points (?) was performed. We'll show the scatter plots of filtered data.
```{r plot frequencies}
LCNEC10_freq_plot <- plot_2d_freq(LCNEC10_clean, K=2)

PANEC1_freq_plot <- plot_2d_freq(PANEC1_clean, K=3)

LCNEC10_freq_plot
PANEC1_freq_plot
```


## Format data

Now we need to format the data to be compatible with the input of Diff() function computing differentiation statistics.
```{r format input}
LCNEC10_subpop_names <- sub(".*\\.", "", colnames(LCNEC10)[3:4])
PANEC1_subpop_names <- sub(".*\\.", "", colnames(PANEC1)[3:5])

LCNEC10_list_freq <- make_popgen_input(LCNEC10_clean[,3:4])
PANEC1_list_freq <- make_popgen_input(PANEC1_clean[,3:5])

# The output of make_popgen_input should be a list of matrices of allele frequencies (2 columns for biallelic loci) for all subpopulations in a sample (in rows)  
LCNEC10_list_freq[[1]]
```
## Compute differentiation statistics

In case of 2 subpopulations one may directly apply Diff() function computing \(M\), \(F_{ST}\), \(G'_{ST}\) and \(D\). Its output should be a list of matrices with \(M\), \(F_{ST}\), \(G'_{ST}\) and \(D\) in rows and their values in the second column. Each matrix correspond to an individual locus from the sample.
```{r compute differentiation statistics for 2 subpops}
LCNEC10_Diff_loci <- lapply(LCNEC10_list_freq, Diff)

LCNEC10_Diff_loci[[1]]

```
In case of 3 subpopulations we may compute the differentiation statistics for the set of all 3 subpopulations as well as for each pair of subpopulations. In this case we need to apply compute_diff_3subpop() function that returns a named list containing four elements: D_123, D_12, D_13 and D_23. Each of the four elements is a list of matrices with differentitation statistics with the same structure as the output of Diff() function. 
```{r compute differentiation statistics for 3 subpops}
PANEC1_Diff_loci = compute_Diff_3subpop(PANEC1_list_freq)

PANEC1_Diff_loci$D_123[[1]]
```
## Plot diferentiation statistics

Once the differentiation statistics have been computed, we may represent them as scatter plots with \(M\) on x-axis and one of the statistics on y-axis, where each point corresponds to a locus. 

```{r}
LCNEC10_combined_plot <- plot_stats(LCNEC10_Diff_loci, K=2, 
                                    title=glue("K=2: {LCNEC10_subpop_names[1]},{LCNEC10_subpop_names[2]}"))


PANEC1_combined_plot <- plot_stats(PANEC1_Diff_loci$D_123, K=3,
                                   title=glue("K=3: {PANEC1_subpop_names[1]},
                                                    {PANEC1_subpop_names[2]}, 
                                                    {PANEC1_subpop_names[3]}"))
LCNEC10_combined_plot
PANEC1_combined_plot
```
  
  Alternatively, for K=3 one can visualize all the computed differentiation statistics: for all subpopulations and for each pair of subpopulations. This can be achieved with plot_stats_3subpop function.

```{r plot stats for 3 subpops}
PANEC1_combined_plot2 <- plot_stats_3subpop(PANEC1_Diff_loci, PANEC1_subpop_names)
```

## Save plots

Save the plots in a folder (that we may want to specify ?)
```{r save plots}
# diff stats plots
#save_plots(LCNEC10_subpop_names, LCNEC10_combined_plot, type="stats", K=2)
#save_plots(PANEC1_subpop_names, PANEC1_combined_plot, type="stats", K=3)

# allele frequency plots
#save_plots(LCNEC10_subpop_names, LCNEC10_freq_plot, type="freq", K=2)
#save_plots(PANEC1_subpop_names, PANEC1_freq_plot, type="freq", K=3)

```


