---
title: "Toad"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Toad}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows an example analysis of yellow toad microsatellite data from Prohl et al. 2021, used to produce Figure 7 from Lim et al.

```{r setup}
library(PopGenBounds)
library(tidyverse)
library(patchwork)
```

We load the dataset
```{r}
Toad_microsat_freq = PopGenBounds::Toad_microsat_freq
toad_genotypes <- PopGenBounds::toad_genotypes
```

# Plotting the allele frequency tables
```{r}
# allele frequency table for the first locus
ggfreqtable(Toad_microsat_freq[[3]])
```

# Computing the statistics
For all subpopulations together (K=47)
```{r}
Diff_toad = lapply(Toad_microsat_freq,Diff)

# check mean statistics values as ratio of averages
HS <- mean(sapply(Diff_toad, function(d) d$value[d$statistic == "HS"]))
HT <- mean(sapply(Diff_toad, function(d) d$value[d$statistic == "HT"]))
1 - HS / HT
```

For pairs of subpopulations (K=2)
```{r}
Diff_toad_pairs = lapply(Toad_microsat_freq,function(f){sapply(1:46, function(i){ sapply((i+1):47, function(j) Diff(f[c(i,j),])$value)})} )
```

# Computing bootstrap confidence intervals
For all subpopulations together (K=47)
```{r}
set.seed(2021)
res <- boot_stat(toad_genotypes, stat_fun = Diff, B = 1000, type = "percentile")
res$mean             # check the labels in the 'statistic' column
```

```{r}
# save bootstrap statistics
per_locus <- do.call(rbind, lapply(names(res$estimate), function(l)
  cbind(locus = l, res$estimate[[l]])))

#avg_limits <- aggregate(cbind(value, value_lower, value_upper) ~ statistic,
#                        data = per_locus, FUN = mean)

# keep the original order of the statistics (M, FST, G'ST, D)
#avg_limits <- avg_limits[match(unique(per_locus$statistic), avg_limits$statistic), ]
#avg_limits

write.csv(per_locus, "../../output/toad_per_locus_basic_ci_hs_ht.csv", row.names = FALSE)
```


# Plotting the data
For K=47
```{r}
toad.tib = tibble(M=unlist( sapply(Diff_toad,function(x){x[1,2]})),
                  FST=unlist( sapply(Diff_toad,function(x){x[2,2]})),
                  GpST=unlist( sapply(Diff_toad,function(x){x[3,2]})),
                  D=unlist( sapply(Diff_toad,function(x){x[4,2]})))

ggtoad = ggbounds_raw(M=toad.tib$M,FST=toad.tib$FST,GpST=toad.tib$GpST,D=toad.tib$D,K=47, ci=res$mean, density_colors = FALSE, ci_style="bands")

ggtoad[[1]]+ ggtoad[[2]]+ ggtoad[[3]]
```


```{r}
fig <- ggtoad[[1]] + ggtoad[[2]] + ggtoad[[3]] + plot_layout(guides = "collect")
ggsave("Toad_files/figures/toad_bounds_CI_bands_ratio_averages.pdf", fig, width = 7, height = 2.3)
```

For K=bp

```{r}
toad.tib_pairs = tibble(M=as.numeric(sapply(Diff_toad_pairs, function(x) unlist(sapply(x,function(y){y[1,]})))),
                  FST=as.numeric(sapply(Diff_toad_pairs, function(x) unlist(sapply(x,function(y){y[2,]})))),
                  GpST=as.numeric(sapply(Diff_toad_pairs, function(x) unlist(sapply(x,function(y){y[3,]})))),
                  D=as.numeric(sapply(Diff_toad_pairs, function(x) unlist(sapply(x,function(y){y[4,]})))) )

# when M=1, no polymorphism, so we exclude the locus
toad.tib_pairs = toad.tib_pairs %>% filter(M<1)

ggtoad_pairs = ggbounds(M=toad.tib_pairs$M,FST=toad.tib_pairs$FST,GpST=toad.tib_pairs$GpST,D=toad.tib_pairs$D,K=2)

ggtoad_pairs[[1]]+ ggtoad_pairs[[2]]+ ggtoad_pairs[[3]]

#ggsave(filename = "Fig_toad.svg",((ggtoad_pairs[[1]]+ scale_color_viridis(limits=c(0,5000)) +guides(color="none")) + 
#  (ggtoad_pairs[[2]]+ scale_color_viridis(limits=c(0,5000)) +guides(color="none") ) + 
#  (ggtoad_pairs[[3]] + scale_color_viridis(limits=c(0,5000)))) / ((ggtoad[[1]]+ scale_color_viridis(limits=c(0,7)) +guides(color="none")) + 
#  (ggtoad[[2]]+ scale_color_viridis(limits=c(0,7)) +guides(color="none") ) + 
#  (ggtoad[[3]] + scale_color_viridis(limits=c(0,7)))),height = 2.3*2,width=3*3)
```
